# G-computation {#sec-g-comp}

{{< include 00-setup.qmd >}}

## The Parametric G-Formula

Let's pause to recap a typical goal of the causal analyses we've seen in this book so far: to estimate what would happen if *everyone* in the study were exposed versus what would happen if *no one* was exposed.
To do this, we've used weighting techniques that create confounder-balanced pseudopopulations which, in turn, give rise to unbiased causal effect estimates in marginal outcome models.
One alternative approach to weighting is called the parametric G-formula, which is generally executed through the following 4 steps:

1. Draw the appropriate time ordered DAG (as described in @sec-dags).

2. For each point in time after baseline, decide on a parametric model that predicts each variable's value based on previously measured variables in the DAG. 
These are often linear models for continuous variables or logistic regressions for binary variables.

3. Starting with the observed distribution of data at baseline, generate values for all subsequent variables according to the models in step 2 (i.e. conduct a *Monte Carlo simulation*). Do this with one key modification: for each exposure regime you are interested in comparing (e.g. everyone exposed versus everyone unexposed), assign the exposure variables accordingly (that is, don't let the simulation assign values for exposure variables).

4. Compute the causal contrast of interest based on the simulated outcome in each exposure group.

::: callout-tip
## Monte Carlo simulations

Monte Carlo simulations are computational approaches that generate a sample of outcomes for random processes.
One example would be to calculate the probability of rolling "snake eyes" (two ones) on a single roll of two six-sided dice. 
We could certainly calculate this probability mathematically ($\frac{1}{6}*\frac{1}{6}=\frac{1}{36}\approx 2.8$%), though it can be just as quick to write a Monte Carlo simulation of the process (1,000,000 rolls shown below).
```{r}
library(tidyverse)

n <- 1000000
tibble(
  roll_1 = sample(1:6, n, replace = TRUE),
  roll_2 = sample(1:6, n, replace = TRUE),
) |> 
  reframe(roll_1 + roll_2 == 2) |> 
  sum()/n
```
Monte Carlo simulations are extremely useful for estimating outcomes of complex processes for which closed mathematical solutions are not easy to determine. 
Indeed, that's why Monte Carlo simulations are so useful for the real-world causal mechanisms described in this book!
:::

## Revisiting the magic morning hours example

Recall in @sec-outcome-model that we estimated the impact of extra magic morning hours on the average posted wait time for the Seven Dwarfs ride between 9 and 10am.
To do so, we fit a propensity score model for the exposure (`park_extra_magic_morning`) with the confounders `park_ticket_season`, `park_close`, and `park_temperature_high`.
In turn, these propensity scores were converted to regression weights for the outcome model, which concluded that the expected impact of having extra magic hours on the average posted wait time between 9 and 10am is 6.2 minutes.

We will now reproduce this analysis, instead adopting the g-formula approach. Proceeding through the 4 steps outlined above, we begin by revisiting the time ordered DAG relevant to this question.

```{r}
#| label: fig-dag-magic-hours-wait-take-2
#| echo: false
#| message: false
#| warning: false
#| fig.cap: >
#|   Proposed DAG for the relationship between Extra Magic Hours
#|   in the morning at a particular park and the average wait
#|   time between 9 am and 10 am.
#|   Here we are saying that we believe 1) Extra Magic Hours impacts average wait time and 2) both Extra Magic Hours and average wait time are determined by the time the park closes, historic high temperatures, and ticket season.

library(ggdag)
library(ggokabeito)

coord_dag <- list(
  x = c(Season = 0, close = 0, weather = -1, x = 1, y = 2),
  y = c(Season = -1, close = 1, weather = 0, x = 0, y = 0)
)

labels <- c(
  x = "Extra Magic Morning",
  y = "Average wait",
  Season = "Ticket Season",
  weather = "Historic high temperature",
  close = "Time park closed"
)

dagify(
  y ~ x + close + Season + weather,
  x ~ weather + close + Season,
  coords = coord_dag,
  labels = labels,
  exposure = "x",
  outcome = "y"
) |>
  tidy_dagitty() |>
  node_status() |>
  ggplot(
    aes(x, y, xend = xend, yend = yend, color = status)
  ) +
  geom_dag_edges_arc(curvature = c(rep(0, 5), .3)) +
  geom_dag_point() +
  geom_dag_label_repel(seed = 1630) +
  scale_color_okabe_ito(na.value = "grey90") +
  theme_dag() +
  theme(
    legend.position = "none",
    axis.text.x = element_text()
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
    limits = c(-1.25, 2.25),
    breaks = c(-1, 0, 1, 2),
    labels = c(
      "\n(one year ago)",
      "\n(6 months ago)",
      "\n(3 months ago)",
      "9am - 10am\n(Today)"
    )
  )
```

The second step is to specify a parametric model for each non-baseline variable that is based upon previously measured variables in the DAG. 
This particular example is simple, since we only have two variables that are affected by previous features (`park_extra_magic_morning` and `wait_minutes_posted_avg`). 
Let's suppose that adequate models for these two variables are the simple logistic and linear models that follow.

```{r}
# Load packages and data
library(broom)
library(touringplans)

seven_dwarfs_9 <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9)

# A logistic regression for park_extra_magic_morning
fit_extra_magic <- glm(
  park_extra_magic_morning ~ 
    park_ticket_season + park_close + park_temperature_high,
  data = seven_dwarfs_9,
  family = "binomial"
)

# A linear model for wait_minutes_posted_avg
fit_wait_minutes <- lm(
  wait_minutes_posted_avg ~ 
    park_extra_magic_morning + park_ticket_season + park_close +
    park_temperature_high,
  data = seven_dwarfs_9
)
```

Next, we need to draw a large sample from the distribution of baseline characteristics.
Deciding how large this sample should be is typically based on computational availability; larger sample sizes can minimize the risk of precision loss via simulation error [@keil2014].
In the present case, we'll use sampling with replacement to generate a data frame of size 10,000.

```{r}
# It's important to set seeds for reproducibility in Monte Carlo runs
set.seed(8675309)

df_sim_baseline <- seven_dwarfs_9 |> 
  select(park_ticket_season, park_close, park_temperature_high) |> 
  sample_n(10000, replace = TRUE)
```

With this population in hand, we can now simulate what would happen at each subsequent time point according to the parametric models we just defined. 
Remember that, in step 3, an important caveat is that for the variable upon which we wish to intervene (in this case, `park_extra_magic_morning`) we don't need to let the model determine the values; rather, we set them. 
Specifically, we'll set the first 5000 to `park_extra_magic_morning = 1` and the second 5000 to `park_extra_magic_morning = 0`. 
Other simulations (in this case, the only remaining variable, `wait_minutes_posted_avg`) proceed as expected.

```{r}
# Set the exposure groups for the causal contrast we wish to estimate
df_sim_time_1 <- df_sim_baseline |> 
  mutate(park_extra_magic_morning = c(rep(1, 5000), rep(0, 5000)))

# Simulate the outcome according to the parametric model in step 2
df_outcome <- fit_wait_minutes |> 
  augment(newdata = df_sim_time_1) |> 
  rename(wait_minutes_posted_avg = .fitted)
```

All that is left to do is compute the causal contrast we wish to estimate.
Here, that contrast is the difference between expected wait minutes on extra magic mornings versus mornings without the extra magic program. 

```{r}
df_outcome |> 
  group_by(park_extra_magic_morning) |> 
  summarize(wait_minutes = mean(wait_minutes_posted_avg))
```

We see that the difference, $74.3-68.1=6.2$ is the same as our estimate of 6.2 when we used IP weighting.

## The g-formula for continuous exposures

As previously mentioned, a key strength of the g-formula is its capacity to handle continuous exposures, a situation in which IP weighting can give rise to unstable estimates. 
Here, we briefly repeat the example from @sec-continuous-exposures to show how this is done.
To extend the pattern, we will wrap this execution of the technique in a bootstrap to show how confidence intervals are computed.

Recall, our causal question of interest is "Do posted wait times for the Seven Dwarfs Mine Train at 8 am affect actual wait times at 9 am?"
The time-ordered DAG for this question (step 1) is:

```{r}
#| label: fig-dag-avg-wait-2
#| echo: false
#| message: false
#| warning: false
#| fig.cap: >
#|    Proposed DAG for the relationship between posted wait
#|    in the morning at a particular park and the average
#|    wait time between 5 pm and 6 pm.

coord_dag <- list(
  x = c(Season = -1, close = -1, weather = -2, extra = 0, x = 1, y = 2),
  y = c(Season = -1, close = 1, weather = 0, extra = 0, x = 0, y = 0)
)

labels <- c(
  extra = "Extra Magic Morning",
  x = "Average posted wait ",
  y = "Average actual wait",
  Season = "Ticket Season",
  weather = "Historic high temperature",
  close = "Time park closed"
)

dagify(
  y ~ x + close + Season + weather + extra,
  x ~ weather + close + Season + extra,
  extra ~ weather + close + Season,
  coords = coord_dag,
  labels = labels,
  exposure = "x",
  outcome = "y"
) |>
  tidy_dagitty() |>
  node_status() |>
  ggplot(
    aes(x, y, xend = xend, yend = yend, color = status)
  ) +
  geom_dag_edges_arc(curvature = c(rep(0, 7), .2, 0, .2, .2, 0), edge_colour = "grey70") +
  geom_dag_point() +
  geom_dag_label_repel(seed = 1602) +
  scale_color_okabe_ito(na.value = "grey90") +
  theme_dag() +
  theme(
    legend.position = "none",
    axis.text.x = element_text()
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(
    limits = c(-2.25, 2.25),
    breaks = c(-2, -1, 0, 1, 2),
    labels = c(
      "\n(one year ago)",
      "\n(6 months ago)",
      "\n(3 months ago)",
      "8am-9am\n(Today)",
      "9am-10am\n(Today)"
    )
  )
```

For step 2, we need to specify parametric models for the non-baseline variables in our DAG (i.e. any variables which have arrows into them).
In this case, we need such models for `park_extra_magic_morning`, `wait_minutes_posted_avg`, and `wait_minutes_actual_avg`; we'll assume that the below logistic and linear models are appropriate.
One extension to our previous implementation is that we're going to embed the each step of the process into a function, since this will allow us to bootstrap the entire pipeline and obtain confidence intervals.

```{r}
fit_models <- function(.data) {
  
  # A logistic regression for park_extra_magic_morning
  fit_extra_magic <- glm(
    park_extra_magic_morning ~ 
      park_ticket_season + park_close + park_temperature_high,
    data = .data,
    family = "binomial"
  )
  
  # A linear model for wait_minutes_posted_avg
  fit_wait_minutes_posted <- lm(
    wait_minutes_posted_avg ~ 
      park_extra_magic_morning + park_ticket_season + park_close +
      park_temperature_high,
    data = .data
  )
  
  # A linear model for wait_minutes_actual_avg
  fit_wait_minutes_actual <- lm(
    wait_minutes_actual_avg ~ 
      wait_minutes_posted_avg + park_extra_magic_morning +
      park_ticket_season + park_close +
      park_temperature_high,
    data = .data
  )
  
  # return a list that we can pipe into our simulation step (up next)
  return(
    list(
      .data = .data,
      fit_extra_magic = fit_extra_magic,
      fit_wait_minutes_posted = fit_wait_minutes_posted,
      fit_wait_minutes_actual = fit_wait_minutes_actual
    )
  )
  
}
```

Next, we write a function which will complete step 3: from a random sample from the distribution of baseline variables, generate values for all subsequent variables (except the intervention variable) according to the models we defined. 

```{r}
# The arguments to simulate_process are as follows:
# fit_obj is a list which is returned from our fit_models function
# contrast gives exposure (default 60) and control group (default 50) settings
# n_sample is the size of the baseline resample of .data
simulate_process <- function(
    fit_obj, 
    contrast = c(60, 50),
    n_sample = 10000
) {
  
  # Draw a random sample of baseline variables
  df_baseline <- fit_obj |> 
    pluck(".data") |> 
    select(park_ticket_season, park_close, park_temperature_high) |> 
    sample_n(n_sample, replace = TRUE)
  
  # Simulate park_extra_magic_morning
  df_sim_time_1 <- fit_obj |> 
    pluck("fit_extra_magic") |> 
    augment(newdata = df_baseline, type.predict = "response") |> 
    # .fitted is the probability that park_extra_magic_morning is 1,
    # so let's use that to generate a 0/1 outcome
    rowwise() |> 
    mutate(
      park_extra_magic_morning = sample(
        x = 1:0,
        size = 1,
        replace = TRUE,
        prob = c(.fitted, 1 - .fitted)
      )
    ) |> 
    ungroup()
  
  # Assign wait_minutes_posted_avg, since it's the intervention
  df_sim_time_2 <- df_sim_time_1 |> 
    mutate(
      wait_minutes_posted_avg = 
        c(rep(contrast[1], n_sample/2), rep(contrast[2], n_sample/2))
    )
  
  # Simulate the outcome 
  df_outcome <- fit_obj |> 
    pluck("fit_wait_minutes_actual") |> 
    augment(newdata = df_sim_time_2) |> 
    rename(wait_minutes_actual_avg = .fitted)
  
  # return a list that we can pipe into the contrast estimation step (up next)
  return(
    list(
      df_outcome = df_outcome,
      contrast = contrast
    )
  )
}
```

Finally, in step 4, we compute the causal contrast of interest using our simulated data. 

```{r}
# sim_obj is a list created by our simulate_process() function
compute_contrast <- function(sim_obj) {
  
  exposure_val <- sim_obj |> 
    pluck("contrast", 1)
  
  control_val <- sim_obj |> 
    pluck("contrast", 2)
  
  sim_obj |> 
    pluck("df_outcome") |> 
    group_by(wait_minutes_posted_avg) |> 
    summarize(avg_wait_actual = mean(wait_minutes_actual_avg)) |> 
    pivot_wider(
      names_from = wait_minutes_posted_avg, 
      values_from = avg_wait_actual,
      names_prefix = "x_"
    ) |> 
    summarize(
      !!sym(paste0("x_", exposure_val)) - !!sym(paste0("x_", control_val))
    )

}
```

Now, let's put it all together to get a single point estimate. 
Once we've seen that in action, we'll bootstrap for a confidence interval.

```{r}
# Wrangle the data to reflect the causal question we are asking
eight <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 8) |>
  select(-wait_minutes_actual_avg)

nine <- seven_dwarfs_train_2018 |>
  filter(wait_hour == 9) |>
  select(park_date, wait_minutes_actual_avg)

wait_times <- eight |>
  left_join(nine, by = "park_date") |>
  drop_na(wait_minutes_actual_avg)

# get a single point estimate to make sure things work as we planned
wait_times |> 
  fit_models() |> 
  simulate_process() |> 
  compute_contrast()
```

## Dynamic treatment regimes with the g-formula


## Calculating estimates with G-Computation

## The Natural Course
