# Target Trials and Standard Methods

{{< include 00-setup.qmd >}}

## Target Trials

## When do standard methods succeed and fail?

When teaching these topics, we are often asked when "standard methods" will succeed, i.e.: when can we just fit a linear regression model to estimate a causal effect? Let's start with the easiest example: the exposure is *randomized*. Why does randomization help? Looking at our assumptions above, traditional randomization solves *consistency* and *well defined exposure* by default. Additionally, it resolves the issue of *exchangeability* because the exposed and unexposed populations (in the limit) are inherently the same since their exposure status was determined by a random process (not by any factors that might make them different from each other). Great! Likewise for *positivity*; if we have randomly assigned folks to either the exposed or unexposed groups, we know the probability of assignment (and we know it is not exactly 0 or 1). Randomization alone does not solve *interference* (for example, if we randomize some people to receive a vaccine for a communicable disease, their receiving it could lower the chance of those around them contracting the infectious disease because it changes the probability of exposure). All that is to say, that randomization ensures comparability and can simplify the methods needed to estimate a causal effect. In the presence of a randomized exposure (assuming perfect adherence to the exposure assigned, no one dropped out of the study, etc.), simple tools like regression can be used to estimate a causal effect. 

### When correlation is causation

When you have no confounders and there is a linear relationship between the exposure and the outcome, that *correlation is a causal relationship*. Even in these cases, using the methods you will learn in this book can help.

1. Adjusting for baseline covariates can make an estimate *more efficient*
1. Propensity score weighting is *more efficient* that direct adjustment
1. Sometimes we are *more comfortable with the functional form of the propensity score* (predicting exposure) than the outcome model

Let's look at an example. I am going to simulate 100 observations. Here the treatment is randomly assigned and there are two baseline covariates: `age` and `weight`. On average, the treatment causes a one unit increase in the outcome (this is called the *average treatment effect*, we will talk more about this quantity in future chapters).

```{r}
#| message: false
#| warning: false
library(tidyverse)
set.seed(10)
n <- 100
d <- tibble(
  age = rnorm(n, 55, 20),
  weight = rnorm(n),
  treatment = rbinom(n, 1, 0.5),
  y = 1 * treatment + 0.2 * age + 0.2 * weight + rnorm(n)
)
```

We can draw a causal diagram of the relationship described in the code above (@fig-diag). Chapter 3 contains more information on these causal diagrams, but briefly, the arrows denote causal relationships, so since we have established that the treatment causes an increase in the outcome (an average treatment effect of 1), we see an arrow from `trt` to `y` in this diagram.

```{r}
#| label: fig-diag
#| echo: false
#| warning: false
#| fig-cap: "Causal Diagram of Example Randomized Study"
library(ggdag)

x <- data.frame(
  name = c("y", "trt", "age", "weight"), 
  time = c(3, 2, 1, 1)
)
dagify(
  y ~ trt + age + weight,
  coords = time_ordered_coords(x)
) |> 
  ggdag() + theme_dag()
```


Let's examine three models: (1) an unadjusted model (@tbl-panel-1), (2) a linear model that adjusts for the baseline covariates (@tbl-panel-2), and (3) a propensity score weighted model (@tbl-panel-3).

::: {#tbl-panel layout-ncol=2}

```{r}
#| code-fold: true
#| message: false
#| warning: false
library(gtsummary)
lm(y ~ treatment, d) |>
  tbl_regression() |>
  modify_column_unhide(column = std.error) |>
  modify_caption("Unadjusted regression")
```

```{r}
#| code-fold: true
#| message: false
#| warning: false
lm(y ~ treatment + age + weight, d) |>
  tbl_regression() |>
  modify_column_unhide(column = std.error) |>
  modify_caption("Adjusted regression")
```


```{r}
#| code-fold: true
#| message: false
#| warning: false
d |>
  mutate(
    p = glm(treatment ~ weight + age, data = d) |> predict(type = "response"),
    ate = treatment / p + (1 - treatment) / (1 - p)
  ) |>
  as.data.frame() -> d
library(PSW)
df <- as.data.frame(d)
x <- psw(df, 
         "treatment ~ weight + age", 
         weight = "ATE", wt = TRUE,
         out.var = "y")
tibble(
  Characteristic = "treatment",
  Beta = round(x$est.wt, 1),
  SE = round(x$std.wt, 3),
  `95% CI` = glue::glue("{round(x$est.wt - 1.96 * x$std.wt, 1)}, {round(x$est.wt + 1.96 * x$std.wt, 1)}"),
  `p-value` = "<0.001"
) |>
  knitr::kable(caption = "Propensity score weighted regression")
```


Three ways to estimate the causal effect.
:::


Looking at the three outputs in @tbl-panel, we can first notice that all three are "unbiased" estimates of the causal effect (we know the true average treatment effect is 1, based on our simulation) -- the estimated causal effect in each table is in the `Beta` column. Great, so all methods give us an unbiased estimate. Next, let's look at the `SE` (standard error) column along with the     `95% CI` (confidence interval) column. Notice the unadjusted model has a *wider* confidence interval (in fact, in this case the confidence interval includes the null, 0) -- this means if we were to use this method, even though we were able to estimate an unbiased causal effect, we would often conclude that we *fail to reject the null* that relationship between the treatment and outcome is 0. In statistical terms, we refer to this as a *lack of efficiency*. Looking at the adjusted analysis in @tbl-panel-2, we see that the standard error is quite a bit smaller (and likewise the confidence interval is tighter, no longer including the null). Even though our baseline covariates `age` and `weight` were not *confounders* adjusting from them *increased the precision* of our result (this is a good thing! We want estimates that are both unbiased *and* precise). Finally, looking at the propensity score weighted estimate we can see that our precision was slightly improved compared to the adjusted result (0.202 compared to 0.204). The magnitude of this improvement will depend on several factors, but it has been shown mathematically that using propensity scores like this to adjust for baseline factors in a randomized trial will *always* improve precision [@williamson2014variance]. What can we learn from this small demonstration? Even in the perfect scenario, where we can estimate unbiased results without using propensity scores, the methods we will show here can be useful. The utility of these methods only increases when exploring more complex examples, such as situations where the effect is *not* randomized, the introduction of time-varying confounders, etc.

What if we did not have a randomized exposure? There are many cases where randomization to a treatment is not ethical or feasible. Standard methods can still estimate unbiased effects, but more care needs to be given to including *all confounders* with their correct functional form. If everything is simple and linear (and there is no effect heterogeneity, that is everyone's causal effect is the same regardless of their baseline factors), then a simple regression model can give you an unbiased result. Let's look at a simple example such as this. Notice in the simulation below, the main difference compared to the above simulation is that the probability of treatment assignment is no longer 0.5 as it was above, but now dependent on the participants `age` and `weight`. The true causal effect is still 1, but now we have two confounders, `age` and `weight` (@fig-diag-2).

```{r}
set.seed(7)
n <- 100000
d <- tibble(
  age = rnorm(n, 55, 20),
  weight = rnorm(n),
  treatment = rbinom(n, 1, 1 / (1 + exp(-0.01 * age - weight))),
  y = 1 * treatment + 0.2 * age + 0.2 * weight + rnorm(n)
)
```

```{r}
#| label: fig-diag-2
#| echo: false
#| warning: false
#| fig-cap: "Causal Diagram of Example Observation Study"

x <- data.frame(
  name = c("y", "trt", "age", "weight"), 
  time = c(3, 2, 1, 1)
)
dagify(
  y ~ trt + age + weight,
  trt ~ age + weight,
  coords = time_ordered_coords(x)
) |> 
  ggdag() + theme_dag()
```

::: {#tbl-panel-2 layout-ncol=2}

```{r}
#| code-fold: true
#| message: false
#| warning: false
lm(y ~ treatment, d) |>
  tbl_regression() |>
  modify_column_unhide(column = std.error) |>
  modify_caption("Unadjusted regression")
```

```{r}
#| code-fold: true
#| message: false
#| warning: false
lm(y ~ treatment + age + weight, d) |>
  tbl_regression() |>
  modify_column_unhide(column = std.error) |>
  modify_caption("Adjusted regression")
```


```{r}
#| code-fold: true
#| message: false
#| warning: false
d |>
  mutate(
    p = glm(treatment ~ weight + age, data = d, family = binomial) |> predict(type = "response"),
    ate = treatment / p + (1 - treatment) / (1 - p)
  ) |>
  as.data.frame() -> d
library(PSW)
df <- as.data.frame(d)
x <- psw(df, 
         "treatment ~ weight + age", 
         weight = "ATE", wt = TRUE,
         out.var = "y")
tibble(
  Characteristic = "treatment",
  Beta = round(x$est.wt, 1),
  SE = round(x$std.wt, 3),
  `95% CI` = glue::glue("{round(x$est.wt - 1.96 * x$std.wt, 1)}, {round(x$est.wt + 1.96 * x$std.wt, 1)}"),
  `p-value` = "<0.001"
) |>
  knitr::kable(caption = "Propensity score weighted regression")
```

Three ways to estimate a causal effect in a non-randomized setting

:::

First, let's look at @tbl-panel-2-1. Here, we see that the unadjusted effect is *biased* (it differs from the true effect, 1, and the true effect is *not* contained in the reported 95% confidence interval). Now lets compare @tbl-panel-2-2 and @tbl-panel-2-3. Technically, both are estimating unbiased causal effects. The output in the `Beta` column of @tbl-panel-2-2 is technically a *conditional* effect (and often in causal inference we want marginal effects), but because there is no treatment heterogeneity in this simulation, the conditional and marginal effects are equal. @tbl-panel-2-3, using the propensity score, also estimates an unbiased effect, but it is no longer the most *efficient* (that was true when the baseline covariates were merely causal for `y`, now that they are `confounders` the efficiency gains for using propensity score weighting are not as clear cut). So why would we ever use propensity scores in this case? Sometimes we have a better understanding of the functional form of the propensity score model compared to the outcome model. Alternatively, sometimes the outcome model is difficult to fit (for example, if the outcome is rare). 

::: callout-tip
## Marginal versus conditional effects

In causal inference, we are often interested in *marginal* effects, mathematically, this means that we want to *marginalize* the effect of interest across the distribution of factors in a particular population that we are trying to estimate a causal effect for. In an adjusted regression model, the coefficients are *conditional*, in other words, when describing the estimated coefficient, we often say something like "a one-unit change in the exposure results in a `coefficient` change in the outcome *holding all other variables in the model constant*. In the case where the outcome is continuous, the effect is linear, and there are no interactions between the exposure effect and other factors about the population, the distinction between an conditional and a marginal effect is largely semantic. If there *is* an interaction in the model, that is, if the exposure has a different impact on the outcome depending on some other factor, we no longer have a single coefficient to interpret. We would want to estimate a *marginal* effect, taking into account the distribution of that factor in the population of interest. Why? We are ultimately trying to determine whether we should suggest the exposure to the target population, so we want to know *on average* whether it will be beneficial. Let's look at quick example: suppose that you are designing an online shopping site. Currently, the "Purchase" button is grey. Changing the button to red increases revenue by \$10 for people who are *not* colorblind and decreases revenue by \$10 for those who *are* colorblind -- *the effect is heterogeneous*. Whether you change the color of the button will depend on the *distribution* of colorblind folks that visit your website. For example, if 50% of the visitors are colorblind, your average effect of changing the color would be \$0. If instead, 100% are colorblind, the average effect of changing the color would be -\$10. Likewise, if 0% are colorblind, the average effect of changing the color to red would be \$10. Your decision, therefore, needs to be based on the *marginal* effect, the effect that takes into account the distribution of colorblind online customers.
:::

<!-- Let's look at an example where the functional form of the propensity score is better understood. Notice in the simulation below, those who have a higher propensity of treatment (those younger than 65 years old), tend to have a wider distribution of age (`sd = 20`). Notice we've also changed the outcome such that there is a non-linear effect of `age` on the outcome that the analyst is not aware of. This does not effect our causal diagram, `age` and `weight` are both still confounders, but it will effect the ability of our adjusted regression model to estimate the correct causal effect. -->

<!-- ```{r} -->
<!-- set.seed(7) -->
<!-- n <- 100000 -->
<!-- d <- tibble( -->
<!--   age = c(rnorm(n/2, 55, sd = 20), rnorm(n/2, 75, sd = 10)), -->
<!--   weight = rnorm(n), -->
<!--   treatment = rbinom(n, 1, 1 / (1 + exp((age > 55) - weight))), -->
<!--   y = 1 * treatment +  -->
<!--     0.2 * age + 0.03 * age^2 +  0.001 * age^3 + # nonlinear effect -->
<!--     0.2 * weight + rnorm(n) -->
<!-- ) -->
<!-- ``` -->

<!-- ::: {#tbl-panel-2 layout-ncol=2} -->

<!-- ```{r} -->
<!-- #| code-fold: true -->
<!-- #| message: false -->
<!-- #| warning: false -->
<!-- lm(y ~ treatment, d) |> -->
<!--   tbl_regression() |> -->
<!--   modify_column_unhide(column = std.error) |> -->
<!--   modify_caption("Unadjusted regression") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| code-fold: true -->
<!-- #| message: false -->
<!-- #| warning: false -->
<!-- lm(y ~ treatment + age + weight, d) |> -->
<!--   tbl_regression() |> -->
<!--   modify_column_unhide(column = std.error) |> -->
<!--   modify_caption("Adjusted regression") -->
<!-- ``` -->


<!-- ```{r} -->
<!-- #| code-fold: true -->
<!-- #| message: false -->
<!-- #| warning: false -->
<!-- d |> -->
<!--   mutate( -->
<!--     p = glm(treatment ~ weight + (age > 55), data = d, family = binomial) |> predict(type = "response"), -->
<!--     ate = treatment / p + (1 - treatment) / (1 - p) -->
<!--   ) |> -->
<!--   as.data.frame() -> d -->
<!-- library(PSW) -->
<!-- df <- as.data.frame(d) -->
<!-- x <- psw(df,  -->
<!--          "treatment ~ weight + age",  -->
<!--          weight = "ATE", wt = TRUE, -->
<!--          out.var = "y") -->
<!-- tibble( -->
<!--   Characteristic = "treatment", -->
<!--   Beta = round(x$est.wt, 1), -->
<!--   SE = round(x$std.wt, 3), -->
<!--   `95% CI` = glue::glue("{round(x$est.wt - 1.96 * x$std.wt, 1)}, {round(x$est.wt + 1.96 * x$std.wt, 1)}"), -->
<!--   `p-value` = "<0.001" -->
<!-- ) |> -->
<!--   knitr::kable(caption = "Propensity score weighted regression") -->
<!-- ``` -->

<!-- Three ways to estimate a causal effect in a non-randomized setting -->

<!-- ::: -->


## Causal inference with `group_by()` and `summarize()`
